### 2020-02-20 Writing/disparity_review.tex
Updated refs

% -Disparity analysis have a history of using ordination techniques
Disparity analyses often use ordination techniques for dimension reduction.
Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
%     -Ordination techniques are blablabla
These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
% -But mainly ordination is good for visualising.
Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
% -Ordination can be good for reducing dimensions if the data has a lot of measurements for a few number of observations (e.g. landmark data, discrete morphological characters, etc.)
Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g. selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012). %TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017}).

@@

Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{(}Bookstein \href{https://paperpile.com/c/sTGYvp/1SD2+sN5d+xaUx+o4w7}{1997, 2015, {[}b{]} 2017, {[}a{]} 2017)}{[}CITE Vera and Thomas paper{]}. %TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \href{https://paperpile.com/c/sTGYvp/0y4V+QVvv}{(Deline et al. 2018; Gerber 2017)}.
Furthermore, interpreting biological variation along the axes is always a \emph{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.

@@

For example, when the analysed data is non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \href{https://paperpile.com/c/sTGYvp/SJbC}{(}Gerber 2014\href{https://paperpile.com/c/sTGYvp/SJbC}{)}. %TG: should be Gerber 2019

@@

Finally, \emph{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
Although some gradients are easy to detect or interpret (e.g. the elongation and depth of mandibles in fishes on first and second PC axes, respectively; {[}\href{https://paperpile.com/c/sTGYvp/3JPy}{Hill et al. 2018{]}}), some are not (e.g. \href{https://paperpile.com/c/sTGYvp/TZzO}{Weisbecker et al. 2019)}.

@@

Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric (\href{https://paperpile.com/c/sTGYvp/SJbC}{i.e. the distance between A and B is not equal to the distance between B and A; Gerber 2014)}.

@@

For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below {[}metrics{]}), it is not necessary to calculate it on ordinated data {[}e.g.\href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}{]}.

@@

For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination (Lloyd 2016, 2018).

@@

\emph{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
% SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
In each case, all three matrices represent the morphospace for the observations at hand.
Visualisation: different ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
% SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
; or ordination axis plots (directly from the ordinated matrix).}
% ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
% SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
% @@@ Change references

\hypertarget{b-disparity-indices}{%
% @@@ Change to indices throughout
\subsection{(b) Measuring disparity}\label{c-disparity-indices}}

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity metric or index \href{https://paperpile.com/c/sTGYvp/vTHS}{(Melanie J. Hopkins and Gerber 2017)}.
As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity {[}metrics paper{]}.
It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g. volume \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, distance from the centroid \href{https://paperpile.com/c/sTGYvp/vTHS+yyNa}{(Melanie J. Hopkins and Gerber 2017; Finlay and Cooper 2015)}, variance and range \href{https://paperpile.com/c/sTGYvp/tGyd}{(S. L. Brusatte et al. 2008)}).
Conversely, if we wish to describe the ``position''
% SG: Note that conceptually and technically,  disparity is about variation. So information on position is another aspect of morphospace occupation. but it is distinct from the meaning of  disparity as originally intended.
 of a group in a morphospace (e.g. does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate (e.g.\href{https://paperpile.com/c/sTGYvp/u1KE}{(Guillerme et al., n.d.)}).
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)}).
%ES: this para is probably most useful to refer to for the ordinate or not part

In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \href{https://paperpile.com/c/sTGYvp/nFf7+ROH8}{(Wills 2001; Ciampaglio, Kemp, and McShea 2001)}.
For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
% SG: Not very clear to me
However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA {[}Legendre \& Legendre 2012{]}.

Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}.
In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \href{https://paperpile.com/c/sTGYvp/Qsl3}{(Bellman 1966)}; Donoho 2000).
Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \href{https://paperpile.com/c/sTGYvp/aSSL}{(}Butler et al. 2012\href{https://paperpile.com/c/sTGYvp/aSSL}{)}
%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
or convex hull based metrics \href{https://paperpile.com/c/sTGYvp/aSSL+PwyQ}{(Butler et al. 2012; Jackson et al. 2011)}.

\hypertarget{c-testing-hypotheses-on-the-evolution-of-disparity}{%
\subsection{(c) Testing hypotheses on the evolution of
disparity}\label{c-testing-hypotheses-on-the-evolution-of-disparity}}

No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer} 2018\href{https://paperpile.com/c/sTGYvp/ZnDd}{)}
%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \href{https://paperpile.com/c/sTGYvp/3hy2+SC6L}{(M. J. Anderson and Walsh 2013; M. J. Anderson 2001)}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
This is not the same as testing whether the two groups overlap in morphospace.
%SG: You're sure of this?
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

@@

%TG: Add section d) within c)
\hypertarget{d-phylogenetic-autocorrelation}{%
\subsection{(d) Phylogenetic
autocorrelation}\label{d-phylogenetic-autocorrelation}}

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \href{https://paperpile.com/c/sTGYvp/WXik}{(Harvey and Pagel 1998)}.
Thus, for many disparity analyses, phylogenetic relationships should be
taken into account.
% SG: Why?
However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
% SG: In what way?
if incorrect assumptions are made about the data (Uyeda et al). \href{https://paperpile.com/c/sTGYvp/ZnDd}{(Adams and Collyer 2018)}
% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
% SG: Which assumption? This part (e) seems a bit vague to me
% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly

%TG: Add section e) within c)???
\hypertarget{e-ancestral-state-estimation-in-disparity-through-time-analyses}{%
\subsection{(e) Ancestral state estimation in disparity-through-time
analyses}\label{e-ancestral-state-estimation-in-disparity-through-time-analyses}}

Disparity-through-time analyses often use ancestral state estimation to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-ordination, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-ordination, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g. the ordination scores) \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.

Pre-ordination ancestral state estimation will change the way the ordination space is defined -- i.e. the relationship between the points are not yet estimated -- and requires longer computational times.
However, once the morphospace is defined its properties will not change.
Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \href{https://paperpile.com/c/sTGYvp/53SJ}{(Lloyd 2018)}.

@@

\hypertarget{disparity-analyses-for-the-future}{%
\section{4. Disparity analyses for the
future}\label{disparity-analyses-for-the-future}}

Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate the bespoke
%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting. There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets \href{https://paperpile.com/c/sTGYvp/krNU+60H0+cV3v}{(Donohue et al. 2013; Canter et al. 2018; Saupe et al. 2015)}. % @@@ Add Mammola
Innovations in morphological disparity analyses likely await discovery in their respective literatures.
% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity

While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
For example, the multidimensional analysis of Diaz \href{https://paperpile.com/c/sTGYvp/47fI}{(Díaz et al. 2016)}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \href{https://paperpile.com/c/sTGYvp/PwyQ}{(Jackson et al. 2011)} can be represented as an isotope-space; ecosystem functioning in Donohue et al. (2013) as an ecosystem-space, etc.
% AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere. Either here as a quick drop, or earlier up under overlap seems appropriate if you want to add it https://esajournals.onlinelibrary.wiley.com/doi/10.1890/14-0235.1
These generalisations could also be exported for any set of traits (e.g. acousto-spaces for acoustic traits or glotto-spaces for linguistic traits).
%SG: devising 
Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
% ES: Suggested by Erin: reword this sentence

>>>>

% -Disparity analysis have a history of using ordination techniques
Disparity analyses often use ordination techniques for dimension reduction.
Dimension reduction (or ordination) is a powerful tool for visualizing data distribution in disparity analyses, helping to elucidate the underlying structure within datasets.
%     -Ordination techniques are blablabla
These are statistical methods to rearrange observed variables so that similar observations are closer together than dissimilar ones (e.g. principal component analysis -- PCA; principal coordinates analysis -- PCO/PCoA; non-metric multidimensional scaling -- NMDS; etc.).
Ordination techniques come in many flavours depending on the data and the desired morphospace properties.
Quantitative (continuous) data can be reduced using PCA, and qualitative data (or mixed data types) can be reduced using PCoA, which is equivalent to metric multidimensional scaling (MDS) or non-metric multidimensional scaling (NMDS) (see Legendre and Legendre 2012,chapter 9, for a detailed explanation of ordination methods and properties).
% -But mainly ordination is good for visualising.
Ordination is advantageous for plotting and visualising data, and can reveal properties of the morphospace not intuitively captured by disparity metrics (see Metrics section below {[}metrics{]}).
% -Ordination can be good for reducing dimensions if the data has a lot of measurements for a few number of observations (e.g. landmark data, discrete morphological characters, etc.)
Using ordinations in disparity analysis makes it easier for researchers to comprehend patterns in just two or three spatial dimensions at a time - or, indeed, just the first few axes of a PCA or a PCO.
Additionally, after ordinating the data, it is possible to reduce further the number of dimensions by focussing on just a subset of axes (e.g. selecting only those axes that describe the majority of the variation) or by fitting the variation to a prescribed number of axes.
In the case of geometric morphometric data, ordination is particularly useful as it conserves the mathematical properties of the data while efficiently reducing the dimensions (Legendre and Legendre 2012). %TG: cite also: Dryden and Mardia "Statistical Shape Analysis book"
This has clear advantages for interpreting the results.
For example, the axes will represent gradients of biological variation (e.g. elongation and flattening of the beak in birds \citep{Cooney2017-ly}.

@@

Unfortunately, dimension reduction can introduce errors of interpretation when these principal axes are unrepresentative of the data more generally \citep{Bookstein1997-dn, Bookstein2015-yy, Bookstein2017-qk, Bookstein2017-gu,Weisbecker2019-kp}.
 %TG: Should be Bookstein 1991 "morphometric tools for landmark data" for the first reference
Visual interpretations of multidimensional data can be particularly misleading, not least since multidimensional spaces are not necessarily Euclidean even when analysing morphometric data \citep{Deline2018-le, Gerber2017-xi}.
Furthermore, interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question; for example, if the first few ordination axes represent the elongation of the beak in birds, but the question is about wing disparity.

@@

For example, when the analysed data is non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging, even if applying an ordination technique is straightforward, \citep{Gerber2017-xi}. %@@@TG: should be Gerber 2019

@@

Finally, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible. %TG: Cite Gerber 2019 10.1111/pala.12407
Although some gradients are easy to detect or interpret (e.g. the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018-ye}), some are not \citep[][e.g.]{Weisbecker2019-kp}.

@@

Such non-Euclidean spaces often have non-intuitive properties, for example, straight lines viewed in bivariate plots of some two selected dimensions are not actually straight and, even less intuitively, distances are non-metric \citep[][i.e. the distance between A and B is not equal to the distance between B and A]{Gerber2014-ol}.

@@

For example, if a metric (e.g. sum or ranges of variance) used to characterise disparity can use all of the data (see Metrics section below metrics%indices
), it is not necessary to calculate it on ordinated data \citep{Close2015-qi}.

@@

For all of these reasons, multidimensional data analyses should not be ordinated automatically, and careful consideration should be given to whether the aim of the study can be achieved without ordination \citep{lloyd2016,lloyd2018}.

@@

\textit{Figure methods: Morphospaces: different mathematical representations of a} \emph{morphospace.
% SG: The morphospace is the original trait space within which lie the specimens. Then one performs an ordination to visualize the space. So I would distinguish the morphospace from the ordination space. See also Mitteroecker and Huttegger (2009)
A trait matrix can be transformed into a distance matrix (e.g. in Close et al., 2015) or an ordinated matrix (e.g. in Brusatte et al.} \emph{2008).
In each case, all three matrices represent the morphospace for the observations at hand.
Visualisation: different ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); ``flattening'' ordination of the data (e.g. using multidimensional scaling} \emph{forcing all the variance to be contained in two dimensions)
% SG: No, the flattening will not force all the variance to be contained in 2D. First because NMDS is non metric and variance is therefore not preserved as such. Second, because there is a stress associated with the flattening that correlates with the degree of distortion imposed on the data to enforce the 2D representation. Some information (variance) is thus necessarily lost.
; or ordination axis plots (directly from the ordinated matrix).}
% ES: I feel like somewhere we should say that in order for the spaces to be Euclidean, the axes must be on same scale. My little bugbear.
% SG response: I think there are more conditions to be Euclidean (inner product defining the  Euclidean norm..). But surely if axes have different scales and units, it will be difficult to define a distance function and the space won't be metric.
% @@@ Change references

% @@@ Change to indices throughout
\subsection{(b) Measuring disparity}

Most disparity datasets are multidimensional and, consequently, a large component of any disparity analysis involves considering how to extract a meaningful (i.e. interpretable) summary of disparity.
This is usually achieved with a disparity metric or index \citep{Hopkins2017-cf}.
As with any summary of multidimensional data, disparity metrics will reflect only some aspects of the morphological variation, never its whole complexity %metrics paper. CITE moms
It is therefore often beneficial to use more than one metric to summarise different aspects of variation, guided by the aim of the study.

When considering only one dimension, disparity metrics can be used to reflect the spread of the distribution (e.g. the range, quantiles or variance) or its central tendency (i.e., mean, median or mode).
Among these metrics, some will have more attractive properties than others, such as sensitivity to outliers. Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if we want to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), metrics related to the spread of the group in the morphospace are most appropriate (e.g. volume
\citealt{Diaz2016-mr}, distance from the centroid \citealt{Hopkins2017-cf, Finlay2015-ft}, variance and range \citealt{Brusatte2008-fx}).
Conversely, if we wish to describe the ``position''
% SG: Note that conceptually and technically,  disparity is about variation. So information on position is another aspect of morphospace occupation. but it is distinct from the meaning of  disparity as originally intended.
 of a group in a morphospace (e.g. does group A occupy the same space as group B?), metrics related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate. %@@@ CITE moms
Finally, if we aim to characterise the density of morphospace occupation (e.g. is group A more closely packed than group B?) metrics related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015-qi}).
%ES: this para is probably most useful to refer to for the ordinate or not part

In addition to considering what properties of disparity metrics should capture, it is important to also consider the mathematical properties of the metrics and their associated caveats \citep{Wills2001-wh, Ciampaglio2001-iz}.
For example, measuring the full sum of the variance in each dimension of the space does not require we add the covariance between the axes in an ordinated space using a PCA.
% SG: Not very clear to me
However, this is not true of other mathematical spaces or when not all dimensions or elements are considered, even in a PCA \citep{Legendre2012-va}.

Furthermore, multidimensional spaces also have some counter-intuitive properties that need to be considered such as the ``curse of dimensionality'' \citep{Bellman1966-mc}.
In spaces with some axis of variance lower than 1, product-based metrics used as proxies of volumes (e.g. product of ranges, hypervolume, hypercube, etc.) can tend towards zero fairly quickly for spaces with even a modest number of dimensions \citep{Bellman1966-mc}. %Donoho 2000
Some other types of metrics are also extremely sensitive to outliers and can be biased easily by sample size, for example range \citep{Butler2012-tr}
%SG: Or even better, twenty years earlier, Foote (1992), who orginally described the problem in the context of disparity and suggested a rarefaction approach.
or convex hull based metrics \citep{Butler2012-tr, Jackson2011-kq}.

\subsection{(c) Testing hypotheses on the evolution of
disparity}

No matter which disparity metrics have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has been greatly expanded in recent years \citep{Adams2018-mg}
%AG: there are others that should be mentioned - mvmorph for example, which is especially important given that there are extremely different approaches to multivariate data in these different packages
, but some of these advances have yet to be implemented in disparity analyses. Instead, hypothesis testing has mostly been confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance \citep{Anderson2001-qb, Anderson2013-zt}, an analysis of variance (ANOVA) of the pairwise distances between different groups.
Although statistically valid, this test is not always directly related to the hypothesis under question.
For example, PERMANOVA tests whether two groups share the same variance/covariance in a ``distance-space''.
% SG: What is meant by distance-sapce and in what way is it distinguished from the spaces discussed above?
This is not the same as testing whether the two groups overlap in morphospace.
%SG: You're sure of this?
Statistical tests should be employed that are tailored to the question at hand, rather than simply following common practices.

@@

%TG: Add section d) within c)
\subsection{(d) Phylogenetic autocorrelation}

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998-xg}.
Thus, for many disparity analyses, phylogenetic relationships should be
taken into account.
% SG: Why?
However, it has been noted that some popular phylogenetic correction methods (like pPCA) can be inappropriate
% SG: In what way?
if incorrect assumptions are made about the data \citep{Uyeda2015}.
% @@@ref: https://academic.oup.com/sysbio/article/64/4/677/1649888
% SG: Which assumption? This part (e) seems a bit vague to me
% AG: as long as you are using most or ideally all of the variance in a dataset, the rotation of pPCA should have little to no impact on disparity analyses. polly et al 2013 i think
provide a thorough review of multivariate phylogenetic comparative methods, and so we do not consider them further here.
% AG: Given this is given quite short treatment, i think if you need room, you could cut it altogether or just have a sentence in another section stating this even more briefly

%TG: Add section e) within c)???
\subsection{(e) Ancestral state estimation in disparity-through-time analyses}

Disparity-through-time analyses often use ancestral state estimation to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-ordination, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-ordination, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g. the ordination scores) \citep{lloyd2018}.

Pre-ordination ancestral state estimation will change the way the ordination space is defined -- i.e. the relationship between the points are not yet estimated -- and requires longer computational times.
However, once the morphospace is defined its properties will not change.
Post-ordination ancestral state estimation will not change the empirical ordination space and is faster to compute, but it will add elements in the space, whose estimated positions can be problematic for statistical tests and evolutionary inferences down the line \citep{lloyd2018}.

@@

\section{4. Disparity analyses for the future}

Morphological disparity analyses are widely employed in evolutionary palaeobiology, and they are based on a diversity of methods and data.
There is no ``one-size-fits-all'' pipeline for morphological disparity analyses.
As with any multidimensional analysis, there are many variables that have to be considered when deciding which data to use and how to analyse it, stemming from the explicit hypotheses being tested.
Many of the problems in morphological disparity analysis arise from ``blind'' application of established methodological pipelines without consideration of the biological question being addressed.
We advocate the bespoke
%VW: I think there could be just a line or two that takes makes "bespoke design" less daunting. There are loads of well-documented R packages available in morphological and ecological studies that can be relatively straightforwardly adapted.
assembly of analytic approaches and, given the computational efficiency of these methods, an experimental approach that explores the impact of competing approaches, such as choice of distance measure, ordination method and ancestral state estimation method on disparity analysis results.
Many of the methods employed in disparity analysis are used more widely in other fields, including genomics and ecology, which also encompass analyses of multidimensional datasets
\citep{Donohue2013-bg, Saupe2015-vm, Canter2018-hk, mammola2019}
. % @@@ Add Mammola
Innovations in morphological disparity analyses likely await discovery in their respective literatures.
% AG: as above, i think you need to add in evodevo work here, it's really fundamental to disparity

While studies of morphological disparity would benefit from advances in multidimensional analysis in other fields, the concept of a morphospace could reciprocally benefit other disciplines.
For example, the multidimensional analysis of Diaz \citealt{Diaz2016-mr}, which analysed patterns of form and function in plants, is essentially an ecomorphospace; isotopic analyses of organisms \citep{Jackson2011-kq} can be represented as an isotope-space; ecosystem functioning in \citealt{Donohue2013-bg} as an ecosystem-space, etc.
% AJ: this is a potential nice, although rather data hungry I believe, methods that I think could be worth dropping into the paper somewhere. Either here as a quick drop, or earlier up under overlap seems appropriate if you want to add it https://esajournals.onlinelibrary.wiley.com/doi/10.1890/14-0235.1
These generalisations could also be exported for any set of traits (e.g. acousto-spaces for acoustic traits or glotto-spaces for linguistic traits).
%SG: devising 
Cognate approaches have been adopted recently in the analysis of single cell comparative transcriptome data \href{https://paperpile.com/c/sTGYvp/856K}{(Sebé-Pedrós et al. 2018)} where interpretation of the resulting transcriptome-spaces would be improved by heeding the concerns we highlight concerning morphospaces.
% ES: Suggested by Erin: reword this sentence
